Viscous Transport in Eroding Porous Media

### 2019-10-03 docs/transport/erosion_transport.tex
NM

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with 20 eroding bodies, but with a smaller
initial porosity.  In Figure~\ref{fig:ErodingLow20vort}, we plot the
eroding geometry and vorticity at four evenly spaced instances in time.
Initially, the smallest distance between pairs of bodies is $3.29 \times
10^{-4}$, and the smallest distance between the bodies and solid wall is
$4.50 \times 10^{-3}$.  At these distances, a resolution of
approximately $N_\iin = 27,000$ and $N_\out = 18,000$ discretization
points is required to satisfy the $5h$ threshold required to guarantee
that the trapezoid rule achieves machine precision.  By applying the
Barycentric quadrature rule, we stably simulate erosion at the much
coarser resolution of $N_\iin = 256$ and $N_\out = 1024$ discretization
points.

@@

To avoid computing streamlines at every time step during the erosion
process, we compute the tortuosity using the Eulerian
method~\eqref{eqn:tortuosity2}.  The initial porosity is $30.67\%$ and
the initial tortuosity is $T = 1.24$.  In
Figure~\ref{fig:ErodingLow20Transport}(a), we plot the tortuosity with
respect to the porosity (blue) and the line of best fit (black) using
the power law 
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.1669}.
\end{align}
This model outperformed the other three models in
equation~\eqref{eqn:tortuosityModels}, and its root-mean-squared error
is $1.13 \times 10^{-2}$.  As grains erode, there is an increase in the
number of streamlines that take a nearly direct pact through the
geometry, and this decreases the tortuosity. However, the channelization
effect of erosion results in an increase in the tortuosity since the
length of many of the streamlines increases when they deflect from a
high porosity region (low pressure) to a low porosity region (high
pressure). For this example, we see that the net effect is a decrease in
the tortuosity for all time.

In Figure~\ref{fig:ErodingLow20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  As in the last
example, we analyze the spreading at several different porosities and we
use the reinsertion algorithm described in Section~\ref{sec:dispersion}
to form sufficiently long trajectories to observe an asymptotic
dispersion rate.  For all the porosities, the dispersion is much closer
to ballistic when compared to the results in
Figure~\ref{fig:Eroding20Transport}.  However, there are still clear
transitions from ballistic dynamics to asymptotic super-dispersive
spreading.  In contrast to the higher porosity initial condition
(Section~\ref{sec:Eroding20}), at early times the erosion results in a
decrease in the dispersion rate. In particular, once erosion has eroded
the first 5\% of the bodies, the particle spreading transitions from
$\sigma_\lambda \sim t^{0.92}$ to $\sigma_\lambda \sim t^{0.69}$.  To
explain this behavior, recall that anomalous dispersion is caused by
tracers spending time in both the fast and slow regimes.  Since the
initial configuration has a reasonably uniform velocity (see
Figure~\ref{fig:ErodingLow20vort}), albeit a small one, the dispersion
is nearly ballistic. However, as the geometry erodes, the flow becomes
more intermittent, and this results in an increased anomalous dispersion
rate~\cite{dea-leb-den-tar-bol-dav2013}.  Then, as the bodies continue
to erode, the geometry channelizes, and most tracers are transported
with a large velocity through the channels, again resulting in a nearly
ballistic motion~\cite{sie-ili-pri-riv-gua2019}.

@@

We compute the tortuosity using both the
Lagrangian~\eqref{eqn:tortuosity1} and Eulerian
methods~\eqref{eqn:tortuosity2}. Therefore, we must compute the velocity
at $N_p = 1000$ points along the inlet $x=-1$. These normalized
velocities are plotted in Figure~\ref{fig:Eroding100tort}(a) for the
eroded geometry at porosity $\phi = 62.98\%$
(Figure~\ref{fig:Eroding100tort}(c)).  The initial velocity of the
tracers is qualitatively similar to the 20 body example
(Figure~\ref{fig:Eroding20tort}(a)), except with additional maximum and
minimums because of the additional bodies.  Next, in
Figure~\ref{fig:Eroding100tort}(b), we plot the local tortuosity by
finding the length of the streamlines as they pass from $x=-1$ to $x=1$.
Compared to Figure~\ref{fig:Eroding20tort}(b), the local tortuosity is
much more discontinuous.  These discontinuities can be explained by
examining the trajectories of tracers in
Figure~\ref{fig:Eroding100tort}(c).  Here, there are many instances of
nearby streamlines that are deflected apart from one another as they
tend to a stagnation point in the flow, and this results in trajectories
with significantly different lengths.  At this porosity, one of the
tracers travels 25.5\% farther than it would have if the bodies had been
absent, and the average tracer travelled 12\% farther than it would have
if the bodies were absent, so the tortuosity is $1.12$.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{20 Bodies at a Low Porosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We consider a second example with 20 eroding bodies, but with a smaller
initial porosity.  In Figure~\ref{fig:ErodingLow20vort}, we plot the
eroding geometry and vorticity at four evenly spaced instances in time.
Initially, the smallest distance between pairs of bodies is $3.29 \times
10^{-4}$, and the smallest distance between the bodies and solid wall is
$4.50 \times 10^{-3}$.  At these distances, a resolution of
approximately $N_\iin = 27,000$ and $N_\out = 18,000$ discretization
points is required to satisfy the $5h$ threshold \nick{\st{required to guarantee
that} needed for} the trapezoid rule to achieve machine precision.  By applying the
Barycentric quadrature rule, we stably simulate erosion at the much
coarser resolution of $N_\iin = 256$ and $N_\out = 1024$ discretization
points.

@@

To avoid computing streamlines at every time step during the erosion
process, we compute the tortuosity using the Eulerian
method~\eqref{eqn:tortuosity2}.  The initial porosity is $30.67\%$ and
the initial tortuosity is $T = 1.24$.  In
Figure~\ref{fig:ErodingLow20Transport}(a), we plot the tortuosity with
respect to the porosity (blue) and the line of best fit (black) using
the power law 
\begin{align}
  \widehat{T}(\phi) = \phi^{-0.1669}.
\end{align}
This model outperformed the other three models in
equation~\eqref{eqn:tortuosityModels}, and its root-mean-squared error
is $1.13 \times 10^{-2}$.  As grains erode, there is an increase in the
number of streamlines that take a nearly direct \nick{\st{pact} path} through the
geometry, and this decreases the tortuosity. However, the channelization
effect of erosion results in an increase in the tortuosity since the
length of many of the streamlines increases when they deflect from a
high porosity region (low pressure) to a low porosity region (high
pressure). For this example, we see that the net effect is a decrease in
the tortuosity for all time.

In Figure~\ref{fig:ErodingLow20Transport}(b), we plot the temporal
evolution of the particle spreading $\sigma_\lambda$.  As in the last
example, we analyze the spreading at several different porosities and we
use the reinsertion algorithm described in Section~\ref{sec:dispersion}
to form sufficiently long trajectories to observe an asymptotic
dispersion rate.  For all the porosities, the dispersion is much closer
to ballistic when compared to the results in
Figure~\ref{fig:Eroding20Transport}.  However, there are still clear
transitions from ballistic dynamics to asymptotic super-dispersive
spreading.  In contrast to the higher porosity initial condition
(Section~\ref{sec:Eroding20}), at early times the erosion results in a
decrease in the dispersion rate. In particular, \nick{once erosion has eroded}
\todo[inline]{Erosion has eroded?}
the first 5\% of the bodies, the particle spreading transitions from
$\sigma_\lambda \sim t^{0.92}$ to $\sigma_\lambda \sim t^{0.69}$.  To
explain this behavior, recall that anomalous dispersion is caused by
tracers spending time in both the fast and slow regimes.  Since the
initial configuration has a reasonably uniform velocity (see
Figure~\ref{fig:ErodingLow20vort}), albeit a small one, the dispersion
is nearly ballistic. However, as the geometry erodes, the flow becomes
more intermittent, and this results in an increased anomalous dispersion
rate~\cite{dea-leb-den-tar-bol-dav2013}.  Then, as the bodies continue
to erode, the geometry channelizes, and most tracers are transported
with a large velocity through the channels, again resulting in a nearly
ballistic motion~\cite{sie-ili-pri-riv-gua2019}.

@@

We compute the tortuosity using both the
Lagrangian~\eqref{eqn:tortuosity1} and Eulerian
methods~\eqref{eqn:tortuosity2}. Therefore, we must compute the velocity
at $N_p = 1000$ points along the inlet $x=-1$. These normalized
velocities are plotted in Figure~\ref{fig:Eroding100tort}(a) for the
eroded geometry at porosity $\phi = 62.98\%$
(Figure~\ref{fig:Eroding100tort}(c)).  The initial velocity of the
tracers is qualitatively similar to the 20 body example
(Figure~\ref{fig:Eroding20tort}(a)), except with additional maximum and
minimums because of the additional bodies.  Next, in
Figure~\ref{fig:Eroding100tort}(b), we plot the local tortuosity by
finding the length of the streamlines as they pass from $x=-1$ to $x=1$.
Compared to Figure~\ref{fig:Eroding20tort}(b), the local tortuosity is
much more discontinuous.  These discontinuities can be explained by
examining the trajectories of tracers in
Figure~\ref{fig:Eroding100tort}(c).  Here, there are many instances of
nearby streamlines that are deflected apart from one another as they
tend to a stagnation point in the flow, and this results in trajectories
with significantly different lengths.  At this porosity, one of the
tracers travels 25.5\% farther than it would have if the bodies had been
absent, and the average tracer travelled 12\% farther \nick{\st{than it would have
if the bodies were absent,} indicating a tortuosity of $1.12$.}
\todo[inline]{Here and in a previous example, I believe the qualifier `than it would have if the bodies were absent' is obvious from the first part of the sentence and does not need to be repeated because it makes the sentence very verbose.}
